---
title: "Exercise 1A"
subtitle: Business Analytics and Data Science WS16/17
output: html_document
---

```{r setup, include=FALSE}
# NEED FOR CHANGE ---- 
# Note that this code will not work on your computer. 
if (!grepl("Dropbox/Lecture/Mod_BADS/Tutorial", getwd(), ignore.case = T)){
  setwd(paste0(getwd(), "/../Dropbox/Lecture/Mod_BADS/Tutorial"))  
}

# You can uncomment the next line and specify the directory for the exercise on your computer
#setwd("C://Your/own/path")
# End of NEED FOR CHANGE ----

solution <- TRUE
if(solution == FALSE){
knitr::opts_chunk$set(echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, fig.keep = 'none', cache = FALSE)
}
options(repos=c(CRAN = "https://cran.uni-muenster.de/"))
```

## A note on R Markdown

This document is written in a format called R Markdown. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see http://rmarkdown.rstudio.com.
It allows us to easily create a pdf file with the exercises, some readable code to give to you, and a nice html file with the solutions and outputs. It will also give you an idea on how to print nice html or pdf reports with R.

After you receive the solutions, you will see that the R code for the solutions are so-called code chunks. When you click the **Knit** button in RStudio ("Knit HTML"), a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. If you want to create a pdf file, you will need to install LaTeX to your computer.

We hope that this will be both easier for you to read in RStudio and easier to save or print.

## Exercise 1.1: Packages and R Studio
A variety of useful R packages are available at CRAN.

1. Navigate to <http://cran.r-project.org/> and look for a package called rpart. Have a brief look into its documentation to learn about its functionality
2. Use R Studio to download and install the package **rpart**
3. Load the package using the **library** command
4. Use the R help to query the functioning of the method **rpart**

```{r, include = FALSE}
#Use R Studio to download and install the package rpart
install.packages("rpart")
#Load the package using the library command
library(rpart)
#Use the R help to query the functioning of the method rpart
help(rpart)
```

## Exercise 1.2: Variables and basic operations  
1. Create two variables $a$ and $b$ and assign values of $3$ and $4.5$.
2. Query the type of variable $a$.
3. Check whether variable $b$ is a text variable (type character).
4. Calculate $a^2 + \frac{1}{b}$, $\sqrt{a*b}$, and $log_2(a)$.

```{r}
#Create two variables a and b and assign values of 3 and 4.5
a <- 3
b <- 4.5
# Check the data type of variable a
typeof(a)
#Check whether variable b is a text variable (type character)
is.character(b)
#Calculate 
a^2+1/b
sqrt(a*b)
log(a,base = 2)
# or alternatively
log2(a)
```

## Exercise 1.3: Variables and basic operations
Create three additional variables as follows:

 $$ A = \left( \begin{matrix} 1 & 2 & 3 \\ 4 & 5 & 6 \\ 7 & 8 & 10 \end{matrix} \right) \quad
  B = \left( \begin{matrix} 1 & 4 & 7 \\ 2 & 5 & 8 \\ 3 & 6 & 9 \end{matrix} \right)  \quad
  y = \left( \begin{matrix} 1 \\ 2 \\ 3 \end{matrix} \right) $$

Calculate  

  1. $a*A$
  2. $A*B$
  3. The inverse of matrix $A$ and store the result in a variable $invA$
  4. Multiply $A$ and $invA$ and verify that the result is the identity matrix
  5. The transpose of matrix $B$
  6. Fill the first row of matrix $B$ with ones
  
```{r}
#Create three additional variables as follows:
A <- matrix(c(1:8,10),nrow = 3, ncol = 3, byrow = TRUE)
B <- matrix(c(1:9),nrow = 3, ncol = 3)
y <- 1:3
#Calculate a*A, A*B
a*A
A%*%B
#The inverse of matrix  A and store the result in a variable  invA
invA <- solve(A)
#verify that the result is the identity matrix
A%*%invA
#The transpose of matrix  B
t(B)
#Fill the first row of matrix B with ones
B[1,] <- c(1,1,1) # or alternatively
B[1,] <- rep(x=1, times=dim(B)[2]) # or even simpler
B[1,] <- 1
```

## Exercise 1.4: Variables and basic operations
1. Check the values of variables $A$, $B$, and $y$ from exercise 1.3
2. Access the second element in the third row of $A$ and the first element in the second row of $B$, and compute their product
3. Multiply the first row of $A$ and the third column of $B$
4. Calculate the least square estimator $\beta$  (we discuss the details in the lecture on regression) $$ \beta = (A^{\top}A)^{-1}A^{\top} y $$

```{r}
#Check the variables. Sidenote: A semicolon is equivalent to a new line
A; B
y
#Access the second element in the third row of A and the first element 
#in the second row of B, and compute their produc
A[3,2] * B[2,1]
#Multiply the first row of A and the third column of B
A[1,]*B[,3]
#Calculate the least square estimator
beta = solve(t(A)*A)%*%y
beta
```

## Exercise 1.5: Using inbuilt functions
The density of the normal distribution with expected value $\mu$ and variance $\sigma$ is given as
$$f(x | \mu ,\sigma ^{2}) = {\frac {1}{\sqrt {2\sigma ^{2}\pi}}}e^{-{\frac {(x-\mu )^{2}}{2\sigma ^{2}}}}$$

1. The function **dnorm** implements the normal distribution in R. Find a way how to compute values of the standard normal distribution using this function.
2. Calculate the value of the standard normal distribution (with $\mu=0$ and $\sigma=1$) at $x=-2, -1.8, \dots, +2$ and store the results in a variable **nvValues**
3. Create a simple graph of the resulting values using the **plot** function

```{r}
x=seq(from=-2, to=2, by=0.2)
#Calculate the value of the standard normal distribution (with mean=0 and sd=1) at x=-2, -1.8, ..., +2 and store the results in a variable nvValues
nvValues=dnorm(x,mean=0,sd=1)
#Create a simple graph of the resulting values using the plot function
plot(x,nvValues,type="l",lwd=2,col="red")
```

## Exercise 1.6: Working with data.frames
1. The UCI Machine Learning repository provides a large number of data sets relating to a variety of domains. Download the data set German Credit from the UCI Repository (use the version german.data as opposed to german.numeric)
2. Load the data into a variable ger using the function **read.csv()**
3. Note that you will have to set some function parameters to read the data properly
4. Use  a text editor for a first examination of the data. Then check the help of **read.csv** and determine which parameters you have to set.
5. Query the type of variable **ger**

```{r}
#Load the data into a variable ger using the function.
# ASSUMPTION: data is available in working directory.
ger <- read.table("german.data",header = FALSE, sep =" " )
#Query the type of variable ger
typeof(ger)
```

## Exercise 1.7: Working with data.frames
The data set does not include column headers. However, a data dictionary is available in the file german.doc on the UCI website. 

1. Read the data dictionary. Then identify an appropriate R function to set the column headers (i.e., variable names) in the data set
2. Set all variable names in the data set and save the resulting data set to a CSV file named **GermanCredit.csv**
3. Use appropriate R functions to answer the following questions
      - What is the most frequent credit purpose? 
      - What are the minimal, maximal and average credit amounts?
      - What is the fraction of good and bad credit risks in the data set?
  
```{r}
#Set all variable names in the data set
colnames(ger) <- c("Status of existing checking account", "Duration","Credit history",
                   "Purpose","Credit amount","Savings account/bonds",
                   "Present employment since","Installment rate","Personal status and sex",
                   "Other debtors/guarantors","Preset residence since",
                   "Property","Age in years","Other installment plans","Housing",
                   "Number of existing credits at this bank","Job"
                   ,"Number of people being liable to provide maintenance for","Telephone",
                   "foreign worker","class")
#Save the resultingdata set to a CSV file named GermanCredit.csv
write.table(ger,"GermanCredit.csv")
#What is the most frequent credit purpose?
names(sort(table(ger[,"Purpose"]), decreasing=TRUE)[1])

#What are the minnimal, maximal and average credit amounts?
min(ger[,"Credit amount"]) 
max(ger[,"Credit amount"])
mean(ger[,"Credit amount"])

#What is the fraction of good and bad credit risks in the data set?
#i)fraction of good risks
sum(ger[,21] == 1)/nrow(ger)
#ii)fraction of bad  risks
sum(ger[,21] == 2)/nrow(ger)

#Note that we can also use the column name, which was specified in the call to 
#the function colnames in line 69. That is, we get the same result when writing  
#the command in the following way, e.g., for good risks
sum(ger$class == 1)/nrow(ger)
```



